The Constant-Q Transform Spectral Envelope Coefficients: A Timbre Feature Designed for Music

### 2021-05-29 IEEE SP Magazine/05-2021 - Draft/article.tex
Finished first part of solution section.

The multidimensional nature of timbre makes it an attribute that is tricky to quantify in terms of one simple characteristic feature \cite{grey1977}. While it is assumed to be independent from pitch and loudness, it is not really feasible to fully disentangle timbre from those qualities, as timbre is inherently dependent on the spectral content of the sound, which is also defined by its pitch and loudness \cite{moore2004}. Researchers in MIR proposed a number of descriptors to characterize one or more aspects of timbre \cite{peeters2011}, but they mostly resort to using the MFCCs when they need one simple timbre feature \cite{mueller2007}. While the MFCCs were shown to be helpful in some MIR tasks, they were initially designed for speech processing applications \cite{mermelstein1976} and are not necessarily adapted to musical data. In particular, they are derived through an old process which makes use of the mel scale, a perceptual scale experimentally designed 80 years ago to approximate the human auditory system's response \cite{stevens1937}. More recently, a number of data-driven approaches attempted to learn some timbral representations from musical data, but generally in terms of implicit embeddings which are tied to a specific trained model \cite{engel2017}, \cite{pons2017}, and not necessarily as explicit and interpretable features such as the MFCCs, which are still usually preferred as the go-to feature to characterize musical timbre by MIR practitioners.

@@

We propose the CQT-SECs, a novel timbre feature that is well-adapted to musical data, pitch-independent, simple to derive, interpretable, and functional. We will show how to derive it from the CQT, a log-frequency transform which matches the Western music scale, by first decomposing it into a pitch-independent spectral envelope and an energy-normalized pitch component, and then extracting a number of timbral coefficients from the spectral envelope.

\subsection{Assumption and Observations}

@@

We then make a first observation that a pitch change in the audio translates to a linear shift in the log-spectrum \cite{brown1991, brown1992}. Assuming that pitch and timbre are independent, this implies that the same musical object at different pitches would have a similar envelope but a shifted pitch component (while two different musical objects at the same pitch would have different envelopes but a similar pitch component). This is summarized in Equation \ref{eq:observation1}.
\begin{equation}
\label{eq:observation1}
\begin{split}
\begin{cases}
X = E * P \\
X' = E' * P' \\
\end{cases}
\Rightarrow E \approx E'
\end{split}
\end{equation}
where $X$, $E$, $P$ and $X'$, $E'$, $P'$ represent the log-spectrum, envelope, and pitch component for a musical object and for a pitch-shifted version of it, respectively.

We make a second observation that the FT of a convolution of two functions is equal to the point-wise product of their FTs, which is also known as the convolution theorem \cite{proakis1995}. This implies that the FT of the log-spectrum would be equal to the point-wise product of the FT of the envelope and the FT of the pitch component. Given the previous observation, this also implies that the FT of the envelope for a musical object and for a pitch-shifted version of it would be equal. This is summarized in equation \ref{eq:observation2}.
\begin{equation}
\label{eq:observation2}
\begin{split}
\begin{cases}
\mathcal{F}(X) = \mathcal{F}(E) \cdot \mathcal{F}(P) \\
\mathcal{F}(X') = \mathcal{F}(E') \cdot \mathcal{F}(P') \\
\end{cases}
\Rightarrow \mathcal{F}(E) \approx \mathcal{F}(E')
\end{split}
\end{equation}
where $\mathcal{F}(.)$ represents the FT function.

We make a third and final observation that the magnitude FT is shift-invariant \cite{proakis1995}. This implies that the FT of the log-spectrum is equal

Equation \ref{eq:observation3}.
\begin{equation}
\label{eq:observation3}
\begin{split}
\begin{cases}
\mathcal{F}(X) = |\mathcal{F}(X)| \cdot e^{j Arg(\mathcal{F}(X))} \\
\mathcal{F}(X') = |\mathcal{F}(X')| \cdot e^{j Arg(\mathcal{F}(X'))} \\
\end{cases}
\Rightarrow |\mathcal{F}(X)| \approx |\mathcal{F}(X')|
\end{split}
\end{equation}
where $|.|$ and $Arg(.)$ represent the modulus and argument, respectively, for a complex array.

Given the ...,

Equation \ref{eq:conclusion}.
\begin{equation}
\label{eq:conclusion}
\begin{split}
&\mathcal{F}(E) \approx |\mathcal{F}(X)| \\
&\Rightarrow E \approx \mathcal{F}^{-1}(|\mathcal{F}(X)|) \\
&\Rightarrow P \approx \mathcal{F}^{-1}(e^{j Arg(\mathcal{F}(X))})
\end{split}
\end{equation}

\subsection{Derivation}

% Can think about the extraction of the envelop as a white normalization of the pitch by using a weighting function that normalizes the signal spectral density by the spectrum magnitude, just like in the GCC-PHAT

%- The envelope component can additionally be refined, along with the pitch component
%- Incidentally, the pitch component can be used to identify the pitch/key

\section{Numerical Example} % or "V - Computational Example"

@@

We have shown that ...

>>>>

The multidimensional nature of timbre makes it an attribute that is tricky to quantify in terms of one simple characteristic feature \cite{grey1977}. While it is assumed to be independent from pitch and loudness, it is not really feasible to fully disentangle timbre from those qualities, as timbre is inherently dependent on the spectral content of the sound, which is also defined by its pitch and loudness \cite{moore2004}. Researchers in MIR proposed a number of descriptors to characterize one or more aspects of timbre \cite{peeters2011}, but they mostly resort to using the MFCCs when they need one simple timbre feature \cite{mueller2007}. While the MFCCs were shown to be helpful in some MIR tasks, they were initially designed for speech processing applications \cite{mermelstein1976} and are not necessarily well-adapted to musical data. In particular, they are derived through an old process which makes use of the mel scale, a perceptual scale experimentally designed 85 years ago to approximate the human auditory system's response \cite{stevens1937}. More recently, a number of data-driven approaches attempted to learn some timbral representations from musical data, but generally in terms of implicit embeddings which are tied to a specific trained model \cite{engel2017}, \cite{pons2017}, and not necessarily as explicit and interpretable features such as the MFCCs, which are still usually preferred as the go-to feature to characterize musical timbre by MIR practitioners.

@@

%We propose the CQT-SECs, a novel timbre feature that is well-adapted to musical data, pitch-independent, simple to derive, interpretable, and functional. We will show how to derive it from the CQT, a log-frequency transform which matches the Western music scale, by first decomposing it into a pitch-independent spectral envelope and an energy-normalized pitch component, and then extracting a number of timbral coefficients from the spectral envelope.

% We will make an assumption, list 3 observations, and give a conclusion...

\subsection{Deconvolution of the CQT}

@@

This convolution process can be thought of a source-filter model \cite{fant1970} which is here not applied in the time domain but the frequency domain, with the source and the filter being the pitch component and the envelope, respectively.

\emph{Observation 1: A pitch change in the audio translates to a linear shift in the log-spectrum \cite{brown1991, brown1992}.}

Assuming that pitch and timbre are independent, this implies that the same musical object at different pitches would have a similar envelope but a shifted pitch component (while two different musical objects at the same pitch would have different envelopes but a similar pitch component). This is summarized in Equation \ref{eq:observation1}, where $X$, $E$, $P$ and $X'$, $E'$, $P'$ represent the log-spectrum, envelope, and pitch component for a musical object and for a pitch-shifted version of it, respectively.
\begin{equation}
\label{eq:observation1}
\begin{split}
\begin{cases}
X = E * P \\
X' = E' * P' \\
\end{cases}
\Rightarrow E \approx E'
\end{split}
\end{equation}

\emph{Observation 2: the FT of a convolution of two functions is equal to the point-wise product of their FTs, which is also known as the convolution theorem \cite{proakis1995}.}

This implies that the FT of the log-spectrum is equal to the point-wise product of the FT of the envelope and the FT of the pitch component. Given the previous observation, this also implies that the FT of the envelope for a musical object and for a pitch-shifted version of it would be equal. This is summarized in Equation \ref{eq:observation2}, where $\mathcal{F}(.)$ represents the FT function.
\begin{equation}
\label{eq:observation2}
\begin{split}
\begin{cases}
\mathcal{F}(X) = \mathcal{F}(E) \cdot \mathcal{F}(P) \\
\mathcal{F}(X') = \mathcal{F}(E') \cdot \mathcal{F}(P') \\
\end{cases}
\Rightarrow \mathcal{F}(E) \approx \mathcal{F}(E')
\end{split}
\end{equation}

\emph{Observation 3: The magnitude FT is shift-invariant \cite{proakis1995}.}

This implies that the magnitude of the FT of the log-spectrum for a musical object and for a pitch-shifted version of it would be equal. This is summarized in Equation \ref{eq:observation3}, where $|.|$ and $Arg(.)$ represent the modulus and argument, respectively, for a complex array. 
\begin{equation}
\label{eq:observation3}
\begin{split}
\begin{cases}
\mathcal{F}(X) = |\mathcal{F}(X)| \cdot e^{j Arg(\mathcal{F}(X))} \\
\mathcal{F}(X') = |\mathcal{F}(X')| \cdot e^{j Arg(\mathcal{F}(X'))} \\
\end{cases}
\Rightarrow |\mathcal{F}(X)| \approx |\mathcal{F}(X')|
\end{split}
\end{equation}

Given the previous observations, we can therefore conclude that the FT of the envelope could be approximated by the magnitude of FT of the log-spectrum, while the FT of the pitch component by the phase component, which finally gives us estimates for the envelope and pitch component after taking their inverse FTs, as shown in Equation \ref{eq:conclusion}, where $\mathcal{F}^{-1}(.)$ represents the inverse FT function.
\begin{equation}
\label{eq:conclusion}
\begin{split}
\Rightarrow 
\begin{cases}
\mathcal{F}(E) \approx |\mathcal{F}(X)| &\Rightarrow E \approx \mathcal{F}^{-1}(|\mathcal{F}(X)|) \\
\mathcal{F}(P) \approx e^{j Arg(\mathcal{F}(X))} &\Rightarrow P \approx \mathcal{F}^{-1}(e^{j Arg(\mathcal{F}(X))}) \\
\end{cases}
\end{split}
\end{equation}

This deconvolution process can also be thought of the normalization of the log-spectrum by the magnitude of its FT (which here corresponds to the FT of the envelope) leading to a sharper log-spectrum (which here corresponds to the pitch component), in the manner of the generalized cross-correlation phase transform (GCC-PHAT) method which aims at normalizing a cross-correlation function by its magnitude spectrum to sharpen the cross-correlation peaks \cite{knapp1976}.

\subsection{Extraction of the Coefficients}

Once the CQT spectrum has been decomposed into its envelope and pitch component, ...

\section{Numerical Examples} % or "V - Computational Example"

\subsection{Analysis on an Example}

%- The envelope component can additionally be refined, along with the pitch component

\subsection{Comparison on a Dataset}

@@

We have shown that we can derive a simple but function timbre feature which is more adapted to musical data ...
